---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ETo: Quick and Easy Calculation of Daily Reference ET

<!-- badges: start -->
[![R-CMD-check](https://github.com/mt-climate-office/ETo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mt-climate-office/ETo/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/mt-climate-office/ETo/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/mt-climate-office/ETo/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

ETo provides basic utility functions for calculating point-based and spatial daily reference ET. While there are many R packages that calculate reference ET (like [Evapotranspiration](https://cran.r-project.org/web/packages/Evapotranspiration/Evapotranspiration.pdf) and [bigleaf](https://cran.r-project.org/web/packages/bigleaf/bigleaf.pdf)), many only provide monthly ET models, and none that we have found support raster inputs out-of-the-box. `ETo` provides a simple interface for calculating daily reference ET and works with `terra::rast` data out-of-the-box. Currently, the Penman-Monteith (FAO-56 model) and Hargreaves methods are implemented. Functions are also provided to calculate all necessary intermediate variables for calculating ETo. The only input variables needed to calculate ETo are daily average temperature, min temperature and max temperature (for Hargreaves), or daily average temperature, relative humidity, solar radiation and windspeed (for Penman-Monteith).

## Installation

You can install the development version of ETo from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mt-climate-office/ETo")
```

## Example
Here, reference ET is calculated using the example dataset given in the [FAO-56 documentation](https://www.fao.org/3/X0490E/x0490e08.htm#chapter%204%20%20%20determination%20of%20eto): 

```{r example, message=FALSE}
library(ETo)

etr_penman_monteith(
  t_max = 21.5, # Daily maximum temperature in degC
  t_min = 12.3, # Daily minimum temperature in degC
  rh_max = 84, # Daily maximum relative humidity (%)
  rh_min = 63, # Daily minimum relative humidity (%)
  srad = 255.4398, # The incoming shortwave radiation in W m^-2
  ws = 2.77778, # Wind speed in meters/second
  lat = 50.8, # Latitude of observation in decimal degrees
  days = 187, # The day of the year
  wind_height = 10, # The height in meters of the wind speed observation
  elev = 100, # The elevation of the observations in meters.
)
```

It also works with vectors:

```{r example2, message=FALSE}
etr_penman_monteith(
  t_max = c(21.5, 23), # Daily maximum temperature in degC
  t_min = c(12.3, 15), # Daily minimum temperature in degC
  rh_max = c(84, 80), # Daily maximum relative humidity (%)
  rh_min = c(63, 50), # Daily minimum relative humidity (%)
  srad = c(255.4398, 325), # The incoming shortwave radiation in W m^-2
  ws = c(2.77778, 3), # Wind speed in meters/second
  lat = 50.8, # Latitude of observation in decimal degrees
  days = c(187, 188), # The day of the year
  wind_height = 10, # The height in meters of the wind speed observation
  elev = 100, # The elevation of the observations in meters.
)
```
And it works with `terra::rast` rasters!

```{r example3, message=FALSE}
r <- etr_penman_monteith(
  t_max = terra::rast(vals=c(21.5, 23), nrows=1, ncols=2), # Daily maximum temperature in degC
  t_min = terra::rast(vals=c(12.3, 15), nrows=1, ncols=2), # Daily minimum temperature in degC
  rh_max = terra::rast(vals=c(84, 80), nrows=1, ncols=2), # Daily maximum relative humidity (%)
  rh_min = terra::rast(vals=c(63, 50), nrows=1, ncols=2), # Daily minimum relative humidity (%)
  srad = terra::rast(vals=c(255.4398, 325), nrows=1, ncols=2), # The incoming shortwave radiation in W m^-2
  ws = terra::rast(vals=c(2.77778, 3), nrows=1, ncols=2), # Wind speed in meters/second
  lat = 50.8, # Latitude of observation in decimal degrees
  days = terra::rast(vals=c(187, 188), nrows=1, ncols=2), # The day of the year
  wind_height = 10, # The height in meters of the wind speed observation
  elev = 100, # The elevation of the observations in meters.
)

terra::plot(r)
```
