---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ETo: Daily Reference ET Calculations for Rasters

<!-- badges: start -->
[![R-CMD-check](https://github.com/mt-climate-office/ETo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mt-climate-office/ETo/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/mt-climate-office/ETo/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/mt-climate-office/ETo/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

ETo provides basic utility functions for calculating point-based and spatial daily reference ET. Currently, the Penman-Monteith and Hargreaves methods are implemented. Functions are also provided to calculate all necessary intermediate variables for calculating ETo. The only input variables needed to calculate ETo are daily average temperature, min temperature and max temperature (for Hargreaves), or daily average temperature, relative humidity, solar radiation and windspeed (for Penman-Monteith).

## Installation

You can install the development version of ETo from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mt-climate-office/ETo")
```

## Example

Below, the [downscaled CMIP6](https://www.nccs.nasa.gov/services/data-collections/land-based-products/nex-gddp-cmip6) data for Montana that are provided with the package are used to calculate and compare different ETo methods:

```{r example, message=FALSE}
library(ETo)

# Load data. Need to read with terra::rast to unpack to a raster.
srad <- terra::rast(srad) |> terra::subset(1:10)
tmean <- terra::rast(tmean) |> terra::subset(1:10) 
# Convert from K to C
tmean <- tmean - 273.15
tmax <- terra::rast(tmax) |> terra::subset(1:10) 
# Convert from K to C
tmax <- tmax - 273.15
tmin <- terra::rast(tmin) |> terra::subset(1:10) 
# Convert from K to C
tmin <- tmin - 273.15
rh <- terra::rast(rh) |> terra::subset(1:10)
ws <- terra::rast(ws) |> terra::subset(1:10)

# Get a raster grid of elevation for the domain.
elev <- get_elev_from_raster(tmean, z = 3)

# Calculate timeseries of Penman Montieth
penman <- calc_etr_spatial(
 tmean = tmean, srad = srad, rh = rh, ws = ws,
 method = "penman", reference = 0.23, elev = elev
)

# Calculate timeseries of Hargreaves
hargreaves <- calc_etr_spatial(
 tmean = tmean, tmax = tmax, tmin = tmin, method = "hargreaves", elev = elev
)

diff <- penman - hargreaves

# Plot the difference (in mm) between the two methods.
terra::plot(diff, range = c(-0.6, 1.5))
```
